### 2023-09-01 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@19136a1ac60e36dcbf1bf724a135b57c30f310f2 🚀

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover
risk,\textsuperscript{2,3,\textbf{Ruiz-Aravena2022Ecological?}} these
approaches oversimplify the relevant interspecific heterogeneity in
immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{4} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{5} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts.\textsuperscript{6} This creates a disconnect
between coevolutionary theory and most existing ecological frameworks
for mapping spillover risk.

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{7} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{8} or
complexes\textsuperscript{9} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{10} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{11} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{12} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

@@

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?}}
Projecting bat and betacoronavirus phylogeny over space
(fig.~\ref{fig:biogeo}), we find support for the idea that bat community
assembly is directly responsible for a global mosaic of viral evolution.
The distinct groupings (represented by different colors, symbolizing
positions in a subspace formed by the first two phylogenetic principal
components) are essentially equivalent between the two groups, and can
be coarsely delineated as (1) south and southeast Asia; (2) east Asia
(including northern China), west Asia, and the Mediterranean coast; (3)
Eurasia above a northing of 40; and (4) Africa and Latin America. In
some cases, this diverges from expectations about coronavirus
biogeography: for example, previous work has rarely flagged India as a
region of interest, but for both bats and betacoronaviruses, the
subcontinent falls into the same regions as the southeast Asian
peninsula (and indeed, the region is home to known bat hosts of multiple
betacoronavirus subgenera, including nobecoviruses, sarbecoviruses, and
merbecoviruses).\textsuperscript{13}

@@

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are broadly consistent at a global
scale; perfect matching between the biogeographic regions would have
indicated that the signal of virus distribution is fully predicted by
bat hosts ranges. Areas for which the biogeographic regions for bats and
betacoronaviruses differ are primarily (i) southeast Asia and southern
China, and (ii) the Arabian peninsula, which are both regions where
zoonotic transmission has been documented (potentially driving a unique
level of viral sampling effort that generates these patterns). These
spatially limited mismatches nonwithstanding, the large level of
congruence may be surprising, given that cross-species transmission may
play a stronger role in coronavirus diversification than
cospeciation---\textsuperscript{2}a property that would theoretically
allow for substantial broad divergence in their biogeography. However,
host jumps at the family level or higher are relatively rare and
significant events in coronavirus evolutionary
history;\textsuperscript{2,14} as a result, the mosaic of
betacoronavirus phylogeography is assembled from a set of overlapping
smaller coevolutionary systems, superimposed in space and filtered by
the importance of different subgroups in local host communities. For
example, the most speciose and cosmopolitan family of bats, the vesper
bats (Vespertilionidae), are considered the primary hosts of the
subgenus \emph{Merbecovirus} (MERS-like viruses);\textsuperscript{13,14}
but in the Americas, where merbecoviruses are the only lineage present,
they have only been found in other bat taxa (\emph{e.g.}, Molossidae,
Phyllostomidae).\textsuperscript{15--18} At the coarsest scale, these
heterogeneities are lost, and betacoronavirus biogeography tracks the
deep rifts in bat evolutionary history---but within broad regions, the
component coevolutionary systems may have very different dynamics.

@@

Overall, these patterns recapitulate the evolutionary history of both
the order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe
bats (Rhinolophidae) include the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}), the group of pandemic threats that have
been of the greatest interest to
researchers\textsuperscript{\textbf{Latinne2020Origins?}} (and so have
been sampled most intensively;).\textsuperscript{25} The hotspots of
host richness and viral diversity in southeast Asia---both of which are
disproportionately high, considering the global landscape of bat species
richness---are almost entirely driven by viral adaptive radiation
through host switching within this clade\textsuperscript{13,21}. In
contrast, the Neotropical hotspot of viral distinctiveness is driven by
isolation by host vicariance. Out of the four main groups of
betacoronaviruses, only merbecoviruses have been found in animals in the
Americas--- an introduction that is generally presumed to be
ancient.\textsuperscript{13,26} While comparatively understudied, New
World merbecoviruses have been found in the ghost-faced bats
(Mormoopidae), Neotropical leaf-nosed bats (Phyllostomidae), and
free-tailed bats (Molossidae).\textsuperscript{15--18} The former two
groups and a clade of the latter are endemic to the Neotropics, while
the explosive adaptive radiations of the phyllostomids are responsible
for the hotspot of bat diversity in the Amazon.\textsuperscript{27}
Together, these clades of New World bats play host to a distinct regime
of betacoronavirus coevolution.

@@

The existence of well-defined cophylogenetic regions suggests that the
bat-betacoronavirus system is spatially fragmented enough to create
divergent coevolutionary trajectories; in turn, this coevolutionary
mosaic may alter the risk of zoonotic emergence. These ideas are,
respectively, supported by the existence of hotspots of viral uniqueness
and the diverse origins of human betacoronaviruses. Together, this
framework points to a predictable relationship between host community
structure and coevolutionary pressure: phylogeographic structure in bat
hosts (and their diverse immune strategies)\textsuperscript{28} creates
a landscape of selective pressure; the trajectory of viruses'
coevolutionary response is, in turn, constrained by their opportunities
for either specialization or diversification through host jumps and
recombination.

@@

\leavevmode\hypertarget{ref-Sanchez2022Strategy}{}%
\CSLLeftMargin{3. }
\CSLRightInline{Sánchez, C. A. \emph{et al.} A strategy to assess
spillover risk of bat SARS-related coronaviruses in southeast asia.
\emph{Nature Communications} \textbf{13}, (2022).}

\leavevmode\hypertarget{ref-Agosta2010How}{}%
\CSLLeftMargin{4. }
\CSLRightInline{Agosta, S. J., Janz, N. \& Brooks, D. R. How specialists
can be generalists: Resolving the {`parasite paradox'} and implications
for emerging infectious disease. \emph{Zoologia (Curitiba)} \textbf{27},
151--162 (2010).}

\leavevmode\hypertarget{ref-Power2004Pathogen}{}%
\CSLLeftMargin{5. }
\CSLRightInline{Power, A. G. \& Mitchell, C. E. Pathogen Spillover in
Disease Epidemics. \emph{The American Naturalist} \textbf{164}, S79--S89
(2004).}

\leavevmode\hypertarget{ref-Geoghegan2017Comparative}{}%
\CSLLeftMargin{6. }
\CSLRightInline{Geoghegan, J. L., Duchêne, S. \& Holmes, E. C.
Comparative analysis estimates the relative frequencies of co-divergence
and cross-species transmission within viral families. \emph{PLOS
Pathogens} \textbf{13}, e1006215 (2017).}

\leavevmode\hypertarget{ref-Thompson2005Geographic}{}%
\CSLLeftMargin{7. }
\CSLRightInline{Thompson, J. N. \emph{The Geographic Mosaic of
Coevolution}. (University Of Chicago Press, 2005).}

\leavevmode\hypertarget{ref-Thompson1994Coevolutionary}{}%
\CSLLeftMargin{8. }
\CSLRightInline{Thompson, J. N. \emph{The Coevolutionary Process}.
(University of Chicago Press, 1994).}

\leavevmode\hypertarget{ref-Janzen1980When}{}%
\CSLLeftMargin{9. }
\CSLRightInline{Janzen, D. H. When is it Coevolution? \emph{Evolution}
\textbf{34}, 611--612 (1980).}

\leavevmode\hypertarget{ref-Price2002Macroevolutionary}{}%
\CSLLeftMargin{10. }
\CSLRightInline{Price, P. W. \emph{Macroevolutionary Theory on
Macroecological Patterns}. (Cambridge University Press, 2002).
doi:\href{https://doi.org/10.1017/CBO9780511615030}{10.1017/CBO9780511615030}.}

\leavevmode\hypertarget{ref-Gomulkiewicz2007Dos}{}%
\CSLLeftMargin{11. }
\CSLRightInline{Gomulkiewicz, R. \emph{et al.} Dos and don'ts of testing
the geographic mosaic theory of coevolution. \emph{Heredity}
\textbf{98}, 249--258 (2007).}

\leavevmode\hypertarget{ref-Thrall2007Coevolution}{}%
\CSLLeftMargin{12. }
\CSLRightInline{Thrall, P. H., Hochberg, M. E., Burdon, J. J. \& Bever,
J. D. Coevolution of symbiotic mutualists and parasites in a community
context. \emph{Trends in Ecology \& Evolution} \textbf{22}, 120--126
(2007).}

\leavevmode\hypertarget{ref-Ruiz-Aravena2022Ecology}{}%
\CSLLeftMargin{13. }
\CSLRightInline{Ruiz-Aravena, M. \emph{et al.} Ecology, evolution and
spillover of coronaviruses from bats. \emph{Nature Reviews Microbiology}
\textbf{20}, 299--314 (2022).}

>>>>

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover risk,\textsuperscript{2--4} these
approaches oversimplify the relevant interspecific heterogeneity in
immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{5} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{6} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts.\textsuperscript{7} This creates a disconnect
between coevolutionary theory and most existing ecological frameworks
for mapping spillover risk.

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{8} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{9} or
complexes\textsuperscript{10} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{11} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{12} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{13} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

@@

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{2,3} Projecting bat and betacoronavirus
phylogeny over space (fig.~\ref{fig:biogeo}), we find support for the
idea that bat community assembly is directly responsible for a global
mosaic of viral evolution. The distinct groupings (represented by
different colors, symbolizing positions in a subspace formed by the
first two phylogenetic principal components) are essentially equivalent
between the two groups, and can be coarsely delineated as (1) south and
southeast Asia; (2) east Asia (including northern China), west Asia, and
the Mediterranean coast; (3) Eurasia above a northing of 40; and (4)
Africa and Latin America. In some cases, this diverges from expectations
about coronavirus biogeography: for example, previous work has rarely
flagged India as a region of interest, but for both bats and
betacoronaviruses, the subcontinent falls into the same regions as the
southeast Asian peninsula (and indeed, the region is home to known bat
hosts of multiple betacoronavirus subgenera, including nobecoviruses,
sarbecoviruses, and merbecoviruses).\textsuperscript{3}

@@

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are broadly consistent at a global
scale; perfect matching between the biogeographic regions would have
indicated that the signal of virus distribution is fully predicted by
bat hosts ranges. Areas for which the biogeographic regions for bats and
betacoronaviruses differ are primarily (i) southeast Asia and southern
China, and (ii) the Arabian peninsula, which are both regions where
zoonotic transmission has been documented (potentially driving a unique
level of viral sampling effort that generates these patterns). These
spatially limited mismatches nonwithstanding, the large level of
congruence may be surprising, given that cross-species transmission may
play a stronger role in coronavirus diversification than
cospeciation---\textsuperscript{2}a property that would theoretically
allow for substantial broad divergence in their biogeography. However,
host jumps at the family level or higher are relatively rare and
significant events in coronavirus evolutionary
history;\textsuperscript{2,14} as a result, the mosaic of
betacoronavirus phylogeography is assembled from a set of overlapping
smaller coevolutionary systems, superimposed in space and filtered by
the importance of different subgroups in local host communities. For
example, the most speciose and cosmopolitan family of bats, the vesper
bats (Vespertilionidae), are considered the primary hosts of the
subgenus \emph{Merbecovirus} (MERS-like viruses);\textsuperscript{3,14}
but in the Americas, where merbecoviruses are the only lineage present,
they have only been found in other bat taxa (\emph{e.g.}, Molossidae,
Phyllostomidae).\textsuperscript{15--18} At the coarsest scale, these
heterogeneities are lost, and betacoronavirus biogeography tracks the
deep rifts in bat evolutionary history---but within broad regions, the
component coevolutionary systems may have very different dynamics.

@@

Overall, these patterns recapitulate the evolutionary history of both
the order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe
bats (Rhinolophidae) include the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}), the group of pandemic threats that have
been of the greatest interest to researchers\textsuperscript{14} (and so
have been sampled most intensively;).\textsuperscript{25} The hotspots
of host richness and viral diversity in southeast Asia---both of which
are disproportionately high, considering the global landscape of bat
species richness---are almost entirely driven by viral adaptive
radiation through host switching within this
clade\textsuperscript{3,21}. In contrast, the Neotropical hotspot of
viral distinctiveness is driven by isolation by host vicariance. Out of
the four main groups of betacoronaviruses, only merbecoviruses have been
found in animals in the Americas--- an introduction that is generally
presumed to be ancient.\textsuperscript{3,26} While comparatively
understudied, New World merbecoviruses have been found in the
ghost-faced bats (Mormoopidae), Neotropical leaf-nosed bats
(Phyllostomidae), and free-tailed bats
(Molossidae).\textsuperscript{15--18} The former two groups and a clade
of the latter are endemic to the Neotropics, while the explosive
adaptive radiations of the phyllostomids are responsible for the hotspot
of bat diversity in the Amazon.\textsuperscript{27} Together, these
clades of New World bats play host to a distinct regime of
betacoronavirus coevolution.

@@

The existence of well-defined cophylogenetic regions suggests that the
bat-betacoronavirus system is spatially fragmented enough to create
divergent coevolutionary trajectories; in turn, this coevolutionary
mosaic may alter the risk of zoonotic emergence. These ideas are,
respectively, supported by the existence of hotspots of viral uniqueness
and the diverse origins of human betacoronaviruses. Together, this
framework points to a predictable relationship between host community
structure and coevolutionary pressure: phylogeographic structure in bat
hosts (and their diverse immune strategies;)\textsuperscript{28} creates
a landscape of selective pressure; the trajectory of viruses'
coevolutionary response is, in turn, constrained by their opportunities
for either specialization or diversification through host jumps and
recombination.

@@

\leavevmode\hypertarget{ref-Ruiz-Aravena2022Ecology}{}%
\CSLLeftMargin{3. }
\CSLRightInline{Ruiz-Aravena, M. \emph{et al.} Ecology, evolution and
spillover of coronaviruses from bats. \emph{Nature Reviews Microbiology}
\textbf{20}, 299--314 (2022).}

\leavevmode\hypertarget{ref-Sanchez2022Strategy}{}%
\CSLLeftMargin{4. }
\CSLRightInline{Sánchez, C. A. \emph{et al.} A strategy to assess
spillover risk of bat SARS-related coronaviruses in southeast asia.
\emph{Nature Communications} \textbf{13}, (2022).}

\leavevmode\hypertarget{ref-Agosta2010How}{}%
\CSLLeftMargin{5. }
\CSLRightInline{Agosta, S. J., Janz, N. \& Brooks, D. R. How specialists
can be generalists: Resolving the {`parasite paradox'} and implications
for emerging infectious disease. \emph{Zoologia (Curitiba)} \textbf{27},
151--162 (2010).}

\leavevmode\hypertarget{ref-Power2004Pathogen}{}%
\CSLLeftMargin{6. }
\CSLRightInline{Power, A. G. \& Mitchell, C. E. Pathogen Spillover in
Disease Epidemics. \emph{The American Naturalist} \textbf{164}, S79--S89
(2004).}

\leavevmode\hypertarget{ref-Geoghegan2017Comparative}{}%
\CSLLeftMargin{7. }
\CSLRightInline{Geoghegan, J. L., Duchêne, S. \& Holmes, E. C.
Comparative analysis estimates the relative frequencies of co-divergence
and cross-species transmission within viral families. \emph{PLOS
Pathogens} \textbf{13}, e1006215 (2017).}

\leavevmode\hypertarget{ref-Thompson2005Geographic}{}%
\CSLLeftMargin{8. }
\CSLRightInline{Thompson, J. N. \emph{The Geographic Mosaic of
Coevolution}. (University Of Chicago Press, 2005).}

\leavevmode\hypertarget{ref-Thompson1994Coevolutionary}{}%
\CSLLeftMargin{9. }
\CSLRightInline{Thompson, J. N. \emph{The Coevolutionary Process}.
(University of Chicago Press, 1994).}

\leavevmode\hypertarget{ref-Janzen1980When}{}%
\CSLLeftMargin{10. }
\CSLRightInline{Janzen, D. H. When is it Coevolution? \emph{Evolution}
\textbf{34}, 611--612 (1980).}

\leavevmode\hypertarget{ref-Price2002Macroevolutionary}{}%
\CSLLeftMargin{11. }
\CSLRightInline{Price, P. W. \emph{Macroevolutionary Theory on
Macroecological Patterns}. (Cambridge University Press, 2002).
doi:\href{https://doi.org/10.1017/CBO9780511615030}{10.1017/CBO9780511615030}.}

\leavevmode\hypertarget{ref-Gomulkiewicz2007Dos}{}%
\CSLLeftMargin{12. }
\CSLRightInline{Gomulkiewicz, R. \emph{et al.} Dos and don'ts of testing
the geographic mosaic theory of coevolution. \emph{Heredity}
\textbf{98}, 249--258 (2007).}

\leavevmode\hypertarget{ref-Thrall2007Coevolution}{}%
\CSLLeftMargin{13. }
\CSLRightInline{Thrall, P. H., Hochberg, M. E., Burdon, J. J. \& Bever,
J. D. Coevolution of symbiotic mutualists and parasites in a community
context. \emph{Trends in Ecology \& Evolution} \textbf{22}, 120--126
(2007).}
